---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.15.0
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# Creating the Relational Database

```{r}
## PROCESS
# widen the table 
# widen every table and clean them up
# (create subsets to try first) add the rest of the data - gender will become an issue
# Additional things to check - if all countries encoded in the same 
# unique countries
# unique continents
# unique genders
# unique years 
# create the primary keys 


## THINGS LEFT TO DO
# Join tables together and check everything is the same as it was before
# general QAs
# spot checks of numbers if it makes sense
# exchange rate table 
# write a lot more about the process
```

## 0 Reading CSVs

```{r}
library(tidyverse)
library(visdat)
```

```{r}
cost <- read.csv("3. cost_of_cigarettes_2020_WHO.csv")
#5. National taxes on a pack of 20 cigarettes.csv
death <- read.csv("6. Prob_dying_diseases.csv")
smoking_prev <- read.csv("8. Age-standardized estimates of current tobacco use, tobacco smoking and cigarette smoking.csv")
life_expectancy <- read.csv("9. Life expectancy WHO.csv")
exchange_rates <- read.csv("exchange_rates.csv")
```

##  1.1 Widening and Cleaning Smoking Prevalance

```{r}
# Getting rid of empty columns
empty_columns <- colSums(is.na(smoking_prev) | smoking_prev == "") == nrow(smoking_prev)
smoking_prev_1 <- smoking_prev[, !empty_columns]
smoking_prev_1
```

```{r}
# Getting rid of redundant columns. We are only interested in the codes and the final values
smoking_prev_2 <- smoking_prev_1 %>%
  select(-ValueType, -Location.type, -Period.type, -IndicatorCode, -ParentLocation, -Location, -Dim1.type, -Dim1, -FactValueNumeric, -FactComments, -Language, -DateModified, -IsLatestYear)
smoking_prev_2
```

```{r}
# Making the table longer
smoking_prev_3 <- smoking_prev_2 %>%
    spread(key = Indicator,
          value = "Value") %>%
    rename(ContinentCode = ParentLocationCode, CountryCode = SpatialDimValueCode, Year = Period, GenderCode = Dim1ValueCode) %>%
    rename("Estimate current cigarette smoking prevalence %" = "Estimate of current cigarette smoking prevalence (%) (age-standardized rate)") %>%
    rename("Estimate current tobacco smoking prevalence %" = "Estimate of current tobacco smoking prevalence (%) (age-standardized rate)") %>%
    rename("Estimate current tobacco use prevalence %" = "Estimate of current tobacco use prevalence (%) (age-standardized rate)")
smoking_prev_3
```

## 1.2 Widening and Cleaning Cost

```{r}
# Getting rid of empty columns
empty_columns <- colSums(is.na(cost) | cost == "") == nrow(cost)
cost_1 <- cost[, !empty_columns]
cost_1 %>% head(2)
```

```{r}
# Getting rid of redundant columns. We are only interested in the codes and the final values
cost_2 <- cost_1 %>%
  select(-ValueType, -Location.type, -Period.type, -IndicatorCode, -ParentLocation, -Location, -FactValueNumeric, -FactValueTranslationID, -Language, -DateModified, -IsLatestYear)
cost_2
```

```{r}
# Making the table longer
cost_3 <- cost_2 %>%
    spread(key = Indicator,
          value = "Value") %>%
    rename(ContinentCode = ParentLocationCode, CountryCode = SpatialDimValueCode, Year = Period)
cost_3
```

## 1.3 Widening and Cleaning Death

```{r}
# Getting rid of empty columns
empty_columns <- colSums(is.na(death) | death == "") == nrow(death)
death_1 <- death[, !empty_columns]
death_1
```

```{r}
# Getting rid of redundant columns. We are only interested in the codes and the final values
death_2 <- death_1 %>%
  select(-IndicatorCode, -ValueType,  -ParentLocation, -Location.type, -Location, -Period.type, -Dim1.type, -Dim1, -Value, -Language, -DateModified, -FactValueNumericLow, -FactValueNumericHigh, -IsLatestYear)
death_2
```

```{r}
# Making the table longer
death_3 <- death_2 %>%
    spread(key = Indicator,
          value = "FactValueNumeric") %>%
    rename(ContinentCode = ParentLocationCode, CountryCode = SpatialDimValueCode, Year = Period, GenderCode = Dim1ValueCode) %>%
    rename("Probability (%) of dying" = "Probability (%) of dying between age 30 and exact age 70 from any of cardiovascular disease, cancer, diabetes, or chronic respiratory disease")
death_3
```


```{r}
# Getting rid of empty columns
empty_columns <- colSums(is.na(life_expectancy) | life_expectancy == "") == nrow(life_expectancy)
life_expectancy_1 <- life_expectancy[, !empty_columns]
life_expectancy_1
```

```{r}
# Getting rid of redundant columns. We are only interested in the codes and the final values
life_expectancy_2 <- life_expectancy_1 %>%
  select(-IndicatorCode, -ValueType,  -ParentLocation, -Location.type, -Location, -Period.type, -Dim1.type, -Dim1, -Value, -Language, -DateModified, -IsLatestYear)
life_expectancy_2
```

```{r}
# Making the table longer
life_expectancy_3 <- life_expectancy_2 %>%
    spread(key = Indicator,
          value = "FactValueNumeric") %>%
    rename(ContinentCode = ParentLocationCode, CountryCode = SpatialDimValueCode, Year = Period, GenderCode = Dim1ValueCode) 
life_expectancy_3
```






## 2.0 Creating Dimensional Tables


## 2.1 Creating Country Dimension Table


In order to create the country dimension table, the unique combinations of country, country code and continent code of each table has to be obtained. Then these combinations need to be checked against each other to ensure that each code relate to the same country. In order to this, the distinct combinations will be joined together to the shared code with full outer joints so that every code can be considered. Then the "get_unique_name" function has been created to check whether the country names are the same, and if not to result "Conflict". 

```{r}
# Getting country, continent code, country code combinations
smoking_countries <- smoking_prev_1 %>% distinct(SpatialDimValueCode, Location, ParentLocationCode) 
cost_countries <- cost_1 %>% distinct(SpatialDimValueCode, Location, ParentLocationCode) 
death_countries <- death_1 %>% distinct(SpatialDimValueCode, Location, ParentLocationCode) 
life_countries <- life_expectancy_1 %>% distinct(SpatialDimValueCode, Location, ParentLocationCode)
```

```{r}
countries <- full_join(smoking_countries, death_countries, by = "SpatialDimValueCode")
countries <- full_join(countries, life_countries, by = "SpatialDimValueCode")
countries <- full_join(countries, cost_countries, by = "SpatialDimValueCode")
countries
```

```{r}
get_unique_name = function(item_1, item_2, item_3, item_4) {
    """Returns item if all items the same, if not gives conflict. If all NA's gives NA"""
    item_og = c(item_1, item_2, item_3, item_4)

    unique_items <- unique(na.omit(item_og))
    num_unique_items <- length(unique_items)

    item <- ifelse(num_unique_items == 1, unique_items, "Conflict")
    return(item)

}
```

```{r}
# Mutating to get the unique country name and continent code and general cleaning
country_dimension_table <- countries %>% 
    rowwise() %>%
    mutate(CountryName = get_unique_name(Location.x, Location.y, Location.x.x, Location.y.y)) %>%
    mutate(ContinentCode = get_unique_name(ParentLocationCode.x, ParentLocationCode.y, ParentLocationCode.x.x, ParentLocationCode.y.y)) %>%
    select(SpatialDimValueCode, CountryName, ContinentCode) %>%
    rename(CountryCode = SpatialDimValueCode)
country_dimension_table
```

```{r}
# There are no conflicts, which means each code relate to a country that is consistent in all tables
country_dimension_table %>% filter(is.na(CountryName) | CountryName == "Conflict" | ContinentCode == "Conflict" | is.na(ContinentCode))
```

## 2.2 Years table

```{r}
# Getting the unique years into a dataframe
smoking_years <- smoking_prev_1$Period %>% unique()
cost_years <- cost_1$Period %>% unique()
death_years <- death_1$Period %>% unique()
life_years <- life_expectancy_1$Period %>% unique()

years_dimensional_table <- data.frame(Years = c(smoking_years, cost_years, death_years, life_years)) %>% unique() %>% arrange(Years)
                                                                                                           
years_dimensional_table
```

## 2.3 Gender

```{r}
# Getting unique combinations of gendercode and gender
smoking_gender <- smoking_prev_1 %>% distinct(Dim1, Dim1ValueCode) 
death_gender <- death_1 %>% distinct(Dim1, Dim1ValueCode) 
life_gender <- life_expectancy_1 %>% distinct(Dim1, Dim1ValueCode)
life_gender
```

```{r}
genders <- full_join(smoking_gender, death_gender, by = "Dim1ValueCode")
genders <- full_join(genders, life_gender, by = "Dim1ValueCode")
genders
```

```{r}
# Mutating to get the unique genders and continent code and general cleaning
gender_dimension_table <- genders %>% 
    rowwise() %>%
    mutate(Gender = get_unique_name(Dim1.x, Dim1.y, Dim1, NA)) %>%
    select(Dim1ValueCode, Gender) %>%
    rename(GenderCode = Dim1ValueCode)
gender_dimension_table
```

## 2.4 Continents


The same process that has been done to countries is done here. 

```{r}
# Getting unique combinations of continent code and continents
smoking_continents <- smoking_prev_1 %>% distinct(ParentLocationCode, ParentLocation) 
cost_continents <- cost_1 %>% distinct(ParentLocationCode, ParentLocation) 
death_continents <- death_1 %>% distinct(ParentLocationCode, ParentLocation) 
life_continents <- life_expectancy_1 %>% distinct(ParentLocationCode, ParentLocation)
```

```{r}
continents <- full_join(smoking_continents, cost_continents, by = "ParentLocationCode")
continents <- full_join(continents, death_continents, by = "ParentLocationCode")
continents <- full_join(continents, life_continents, by = "ParentLocationCode")
continents
```

```{r}
# Mutating to get the unique continents and general cleaning
continent_dimensional_table <- continents %>% 
    rowwise() %>%
    mutate(ContinentName = get_unique_name(ParentLocation.x, ParentLocation.y, ParentLocation.x.x, ParentLocation.y.y)) %>%
    select(ParentLocationCode, ContinentName) %>%
    rename(ContinentCode = ParentLocationCode)
continent_dimensional_table
```


## 2.5 Exchange Rates

```{r}
# Select only the columns we need 
exchange_rates <- exchange_rates %>%
  select(Country, Currency, Calendar.Year, Exchange.Rate)
exchange_rates

# group by country and year, average the exchange rate
exchange_rates_grouped <- exchange_rates %>%
  group_by(Country, Calendar.Year)  %>%
  summarize(average_fx = round(mean(Exchange.Rate), 2))
exchange_rates_grouped 

```
```{r}
# add country code
exchange_rates_grouped  <- exchange_rates_grouped  %>% mutate(Country = tolower(Country))
colnames(exchange_rates_grouped)[colnames(exchange_rates_grouped) == "Country"] <- "Location.x"

exchange_rates_grouped <- exchange_rates_grouped %>%
  left_join(
    countries %>% select(SpatialDimValueCode, Location.x) %>%
    mutate(Location.x = tolower(Location.x)),
    by = "Location.x")
exchange_rates_grouped

```


## 3.0 Making Fact Tables - Subsets (Trial)


The smoking prevalence, life expectancy and probability of death tables will be combined to one fact table. The cost table will be kept separate as it doesn't have gender. For the trial, a combination that has a value in each table has been chosen. In this case, continent in Africa, ZWE as country and both sexes as gender. This way we can check that the values from each table connects properly. As all the values should be in one row for this specific combination. 

```{r}
# Creating the subset for the specific combination (Contient = AFR, Country = ZWE, Gender = BTSX)
smoking_prev_3_subset <- smoking_prev_3 %>% filter((ContinentCode == "AFR") & (CountryCode == "ZWE") & (GenderCode == "BTSX") & (Year == 2010))
life_expectancy_3_subset <- life_expectancy_3 %>% filter((ContinentCode == "AFR") & (CountryCode == "ZWE") & (GenderCode == "BTSX") & (Year == 2010))
death_3_subset <- death_3 %>% filter((ContinentCode == "AFR") & (CountryCode == "ZWE") & (GenderCode == "BTSX") & (Year == 2010))
```

```{r}
smoking_prev_3_subset
life_expectancy_3_subset
death_3_subset
```

```{r}
trial_2 <- full_join(smoking_prev_3_subset, life_expectancy_3_subset, by = c("ContinentCode", "CountryCode", "Year", "GenderCode"))
trial_3 <- full_join(trial_2, death_3_subset, by = c("ContinentCode", "CountryCode", "Year", "GenderCode"))
trial_3

## it works - all in one row 
```

## 3.1 Joining Tables - Full set

```{r}
smoke_lifexpect <- full_join(smoking_prev_3, life_expectancy_3, by = c("ContinentCode", "CountryCode", "Year", "GenderCode"))
smoke_lifexpect_death <- full_join(smoke_lifexpect, death_3, by = c("ContinentCode", "CountryCode", "Year", "GenderCode"))
smoke_lifexpect_death
```

## 4.0 Creating Primary Keys


We are going to keep the cost table separate to the smoke_lifexpect_death table as the cost table does not have a gender. 


### 4.1 Table 1 - Costs of tobacco

```{r}
cost_3
```

```{r}
cost_cigs <- cost_3 %>%
  mutate(CostID = paste(CountryCode, Year),
         CostID = str_replace_all(CostID, " ", "")) %>%
  select(CostID, everything(), -ContinentCode, -CountryCode, -Year)

cost_cigs
```

### 4.2 Table 2 - Smoking Prevalence, Life Expectancy, Probablity of Death

```{r}
smoke_lifexpect_death_wID <- smoke_lifexpect_death %>%
  mutate(ID = paste(CountryCode, Year, substr(GenderCode, 1, 1)),
         ID = str_replace_all(ID, " ", "")) %>%
  select(ID, everything(), -ContinentCode, -CountryCode, -Year, -GenderCode)

smoke_lifexpect_death_wID
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
