---
title: "DATA 422 Group Project - Wrangling the Data"
output: html_document
date: "2023-10-21"
---

# Creating the Relational Database


The main of the data wrangling is to put the data into a relational database that is easy and clear to use. We wanted one main fact table that had all the variables of interest in each column and smaller dimension tables that held unique values for particular categories. 

The datasets were loaded and it was identified that all the datasets were in long format. Therefore, the datasets were widened to see the variables in each dataset, identify the keys and understand how to join the datasets.

Afterwards, the datasets were joined together to make the final fact table with primary keys based on country, year and gender. The cost and taxes datasets were put under 'Both Sex' category. Dimension tables were created as well for countries, continents, exchange rates, years and gender. This would avoid a repetition of this data in the fact table. 


## 0 Reading CSVs

```{r}
library(tidyverse)
library(visdat)
```

```{r}
cost <- read.csv("3. cost_of_cigarettes_2020_WHO.csv")
taxes <- read.csv("5. National taxes on a pack of 20 cigarettes.csv")
death <- read.csv("6. Prob_dying_diseases.csv")
smoking_prev <- read.csv("8. Age-standardized estimates of current tobacco use, tobacco smoking and cigarette smoking.csv")
life_expectancy <- read.csv("9. Life expectancy WHO.csv")
exchange_rates <- read.csv("7. exchange_rates.csv")
```

##  1.1 Widening and Cleaning Smoking Prevalance

```{r}
# Getting rid of empty columns
empty_columns <- colSums(is.na(smoking_prev) | smoking_prev == "") == nrow(smoking_prev)
smoking_prev_1 <- smoking_prev[, !empty_columns]
smoking_prev_1
```

```{r}
# Getting rid of redundant columns. We are only interested in the codes and the final values
smoking_prev_2 <- smoking_prev_1 %>%
  select(-ValueType, -Location.type, -Period.type, -Indicator, -ParentLocation, -Location, -Dim1.type, -Dim1, -FactValueNumeric, -FactComments, -Language, -DateModified, -IsLatestYear)
smoking_prev_2

```

```{r}
smoking_prev_2 %>% 
    filter(is.na(SpatialDimValueCode) | is.na(Period) | is.na(Dim1ValueCode))
```

```{r}
# Making the table wider
smoking_prev_3 <- smoking_prev_2 %>%
    spread(key = IndicatorCode,
          value = "Value") %>%
    rename(ContinentCode = ParentLocationCode, CountryCode = SpatialDimValueCode, Year = Period, GenderCode = Dim1ValueCode)
smoking_prev_3
```

## 1.2 Widening and Cleaning Cost

```{r}
# Getting rid of empty columns
empty_columns <- colSums(is.na(cost) | cost == "") == nrow(cost)
cost_1 <- cost[, !empty_columns]
cost_1 %>% head(2)
```

```{r}
# Getting rid of redundant columns. We are only interested in the codes and the final values
cost_2 <- cost_1 %>%
  select(-ValueType, -Location.type, -Period.type, -Indicator, -ParentLocation, -Location, -FactValueNumeric, -FactValueTranslationID, -Language, -DateModified, -IsLatestYear)
cost_2
```

```{r}
cost_2 %>% 
    filter(is.na(SpatialDimValueCode) | is.na(Period))
```

```{r}
# Making the table wider
cost_3 <- cost_2 %>%
    spread(key = IndicatorCode,
          value = "Value") %>%
    rename(ContinentCode = ParentLocationCode, CountryCode = SpatialDimValueCode, Year = Period)
cost_3
```

## 1.3 Widening and Cleaning Death

```{r}
# Getting rid of empty columns
empty_columns <- colSums(is.na(death) | death == "") == nrow(death)
death_1 <- death[, !empty_columns]
death_1
```

```{r}
# Getting rid of redundant columns. We are only interested in the codes and the final values
death_2 <- death_1 %>%
  select(-Indicator, -ValueType,  -ParentLocation, -Location.type, -Location, -Period.type, -Dim1.type, -Dim1, -Value, -Language, -DateModified, -FactValueNumericLow, -FactValueNumericHigh, -IsLatestYear)
death_2
```

```{r}
death_2 %>% 
    filter(is.na(SpatialDimValueCode) | is.na(Period) | is.na(Dim1ValueCode))
```

```{r}
# Making the table wider
death_3 <- death_2 %>%
    spread(key = IndicatorCode,
          value = "FactValueNumeric") %>%
    rename(ContinentCode = ParentLocationCode, CountryCode = SpatialDimValueCode, Year = Period, GenderCode = Dim1ValueCode) 
death_3
```

## 1.4 Widening and Cleaning Life Expectancy

```{r}
life_expectancy
```

```{r}
# Getting rid of empty columns
empty_columns <- colSums(is.na(life_expectancy) | life_expectancy == "") == nrow(life_expectancy)
life_expectancy_1 <- life_expectancy[, !empty_columns]
life_expectancy_1
```

```{r}
# Getting rid of redundant columns. We are only interested in the codes and the final values
life_expectancy_2 <- life_expectancy_1 %>%
  select(-Indicator, -ValueType,  -ParentLocation, -Location.type, -Location, -Period.type, -Dim1.type, -Dim1, -Value, -Language, -DateModified, -IsLatestYear)
life_expectancy_2
```

```{r}
life_expectancy_2 %>% 
    filter(is.na(SpatialDimValueCode) | is.na(Period) | is.na(Dim1ValueCode))
```

```{r}
# Making the table wider
life_expectancy_3 <- life_expectancy_2 %>%
    spread(key = IndicatorCode,
          value = "FactValueNumeric") %>%
    rename(ContinentCode = ParentLocationCode, CountryCode = SpatialDimValueCode, Year = Period, GenderCode = Dim1ValueCode) 
life_expectancy_3
```

## 1.5 Widening and Cleaning Taxes

```{r}
# Getting rid of empty columns
empty_columns <- colSums(is.na(taxes) | taxes == "") == nrow(taxes)
taxes_1 <- taxes[, !empty_columns]
```

```{r}
taxes_2 <- taxes_1 %>%
    select(-ValueType, -Location.type, -Period.type, -Indicator, -ParentLocation, -Location, -FactValueNumeric, -Language, -DateModified, -IsLatestYear, -FactValueTranslationID)
taxes_2
```

```{r}
taxes_2 %>% 
    filter(is.na(SpatialDimValueCode) | is.na(Period))
```

```{r}
# Making the table wider
taxes_3 <- taxes_2 %>%
    spread(key = IndicatorCode,
          value = "Value") %>%
    rename(ContinentCode = ParentLocationCode, CountryCode = SpatialDimValueCode, Year = Period) 
taxes_3
```

## 2.0 Creating Dimensional Tables


Through widening the datasets above, it was identified that smoking prevalance, mortality and life expectancy datasets have country, year and gender as the keys while cost and taxes have country and year. Each country is contained in a continent as well and each of the costs given is in its own currency.

Dimension tables will be created for country, year and gender as these will be used to make the primary keys and it will be useful to have them in separatable tables with unique values. Continents will also be put in a dimension table and connected to the country dimension table in order to avoid repeating it in the fact table. An exchange rate dimension table will also be created which can be joined to the fact table to find the cost of tobacco in each country. 


## 2.1 Creating Country Dimension Table


In order to create the country dimension table, the unique combinations of country, country code and continent code of each table has to be obtained. Then these combinations need to be checked against each other to ensure that each code relate to the same country. In order to this, the distinct combinations will be joined together to the shared code with full outer joints so that every code can be considered. Then the "get_unique_name" function has been created to check whether the country names are the same, and if not to result "Conflict". 

```{r}
# Getting country, continent code, country code combinations
smoking_countries <- smoking_prev_1 %>% distinct(SpatialDimValueCode, Location, ParentLocationCode) 
cost_countries <- cost_1 %>% distinct(SpatialDimValueCode, Location, ParentLocationCode) 
death_countries <- death_1 %>% distinct(SpatialDimValueCode, Location, ParentLocationCode) 
life_countries <- life_expectancy_1 %>% distinct(SpatialDimValueCode, Location, ParentLocationCode)
taxes_countries <- taxes_1 %>% distinct(SpatialDimValueCode, Location, ParentLocationCode)
```

```{r}
countries <- full_join(smoking_countries, death_countries, by = "SpatialDimValueCode")
countries <- full_join(countries, life_countries, by = "SpatialDimValueCode")
countries <- full_join(countries, cost_countries, by = "SpatialDimValueCode")
countries <- full_join(countries, taxes_countries, by = "SpatialDimValueCode")
countries
```

```{r}
get_unique_name = function(item_1, item_2, item_3, item_4, item_5) {
   # """Returns item if all items the same, if not gives conflict. If all NA's gives NA"""
    item_og = c(item_1, item_2, item_3, item_4, item_5)

    unique_items <- unique(na.omit(item_og))
    num_unique_items <- length(unique_items)

    item <- ifelse(num_unique_items == 1, unique_items, "Conflict")
    return(item)

}
```

```{r}
# Mutating to get the unique country name and continent code and general cleaning
country_dimension_table <- countries %>% 
    rowwise() %>%
    mutate(CountryName = get_unique_name(Location.x, Location.y, Location.x.x, Location.y.y, Location)) %>%
    mutate(ContinentCode = get_unique_name(ParentLocationCode.x, ParentLocationCode.y, ParentLocationCode.x.x, ParentLocationCode.y.y, ParentLocationCode)) %>%
    select(SpatialDimValueCode, CountryName, ContinentCode) %>%
    rename(CountryCode = SpatialDimValueCode)
country_dimension_table
```

```{r}
# There are no conflicts, which means each code relate to a country that is consistent in all tables
country_dimension_table %>% filter(is.na(CountryName) | CountryName == "Conflict" | ContinentCode == "Conflict" | is.na(ContinentCode))
```

```{r}
duplicates <- country_dimension_table %>%
    group_by(CountryName) %>%
    summarise(count = n()) %>%
    filter(count > 1)
duplicates 
# No duplicates - OK
```

## 2.2 Years table

```{r}
# Getting the unique years into a dataframe
smoking_years <- smoking_prev_1$Period %>% unique()
cost_years <- cost_1$Period %>% unique()
death_years <- death_1$Period %>% unique()
life_years <- life_expectancy_1$Period %>% unique()
taxes_years <- taxes_1$Period %>% unique()

years_dimensional_table <- data.frame(Years = c(smoking_years, cost_years, death_years, life_years, taxes_years)) %>% unique() %>% arrange(Years)
                                                                                                           
years_dimensional_table
```

## 2.3 Gender

```{r}
# Getting unique combinations of gendercode and gender
smoking_gender <- smoking_prev_1 %>% distinct(Dim1, Dim1ValueCode) 
death_gender <- death_1 %>% distinct(Dim1, Dim1ValueCode) 
life_gender <- life_expectancy_1 %>% distinct(Dim1, Dim1ValueCode)
life_gender
```

```{r}
genders <- full_join(smoking_gender, death_gender, by = "Dim1ValueCode")
genders <- full_join(genders, life_gender, by = "Dim1ValueCode")
genders
```

```{r}
# Mutating to get the unique genders and continent code and general cleaning
gender_dimension_table <- genders %>% 
    rowwise() %>%
    mutate(Gender = get_unique_name(Dim1.x, Dim1.y, Dim1, NA, NA)) %>%
    select(Dim1ValueCode, Gender) %>%
    rename(GenderCode = Dim1ValueCode)
gender_dimension_table
```

## 2.4 Continents


The same process that has been done to countries is done here. 

```{r}
# Getting unique combinations of continent code and continents
smoking_continents <- smoking_prev_1 %>% distinct(ParentLocationCode, ParentLocation) 
cost_continents <- cost_1 %>% distinct(ParentLocationCode, ParentLocation) 
death_continents <- death_1 %>% distinct(ParentLocationCode, ParentLocation) 
life_continents <- life_expectancy_1 %>% distinct(ParentLocationCode, ParentLocation)
taxes_continents <- taxes_1 %>% distinct(ParentLocationCode, ParentLocation)
```

```{r}
continents <- full_join(smoking_continents, cost_continents, by = "ParentLocationCode")
continents <- full_join(continents, death_continents, by = "ParentLocationCode")
continents <- full_join(continents, life_continents, by = "ParentLocationCode")
continents <- full_join(continents, taxes_continents, by = "ParentLocationCode")
continents
```

```{r}
# Mutating to get the unique continents and general cleaning
continent_dimensional_table <- continents %>% 
    rowwise() %>%
    mutate(ContinentName = get_unique_name(ParentLocation.x, ParentLocation.y, ParentLocation.x.x, ParentLocation.y.y, ParentLocation)) %>%
    select(ParentLocationCode, ContinentName) %>%
    rename(ContinentCode = ParentLocationCode)
continent_dimensional_table
```

## 2.5 Indicators


These have been obtained just as a reference for later analysis.

```{r}
# Getting the unique years into a dataframe
smoking_indicators <- smoking_prev_1 %>%
  select(Indicator, IndicatorCode) %>%  
  distinct()
cost_indicators <- cost_1 %>%
  select(Indicator, IndicatorCode) %>%  
  distinct()
death_indicators <- death_1 %>%
  select(Indicator, IndicatorCode) %>%  
  distinct()
tax_indicators <- taxes_1 %>%
  select(Indicator, IndicatorCode) %>%  
  distinct() 
life_indicators <- life_expectancy_1 %>%
  select(Indicator, IndicatorCode) %>%  
  distinct() 


indicators <- bind_rows(smoking_indicators, cost_indicators, death_indicators, tax_indicators, life_indicators) %>% distinct()                         
                                                                                                           
indicators
```

## 2.6 Exchange Rates Dimensional Table

```{r}
# Select only the columns we need 
exchange_rates <- exchange_rates %>%
  select(Country, Currency, Calendar.Year, Exchange.Rate)

# group by country and year, average the exchange rate
exchange_rates_grouped <- exchange_rates %>%
  group_by(Country, Calendar.Year)  %>%
  summarize(average_fx = round(mean(Exchange.Rate), 2))

# rename country to CountryaName, lowercase country names
exchange_rates_grouped  <- exchange_rates_grouped  %>% mutate(Country = tolower(Country))
colnames(exchange_rates_grouped)[colnames(exchange_rates_grouped) == "Country"] <- "CountryName"

# join tables and add country code
exchange_rates_grouped <- exchange_rates_grouped %>%
  left_join(
    country_dimension_table %>% select(CountryCode, CountryName) %>%
    mutate(CountryName = tolower(CountryName)),
    by = "CountryName")

# check which country names did not match and count
empty_country_codes <- exchange_rates_grouped %>%
  filter(is.na(CountryCode) | CountryCode == "") %>% 
  select(CountryName)  %>% 
  group_by(CountryName) %>%
  summarize(EmptyEntriesCount = n())
empty_country_codes
```

```{r}
# get distinct counties as defined by WHO 
distinct_countries_WHO = country_dimension_table %>% distinct(CountryName)  %>%  arrange(CountryName)

 # export data and match manually 
#write.csv(distinct_countries_WHO, file = 'who-countries.csv', row.names = FALSE)
#write.csv(empty_country_codes, file = 'fx-countries.csv', row.names = FALSE)

# read in manually matched countries
country_match = read.csv('fx-countries.csv')

# clean up the empty values
exchange_rates_grouped$CountryName = ifelse(exchange_rates_grouped$CountryName %in% country_match$CountryNameFX,    
                                        country_match$CountryNameWHO[match(exchange_rates_grouped$CountryName, country_match$CountryNameFX)],
                                        exchange_rates_grouped$CountryName)
exchange_rates_grouped
```

```{r}
# lowercase country names
exchange_rates_grouped  <- exchange_rates_grouped  %>% mutate(CountryName = tolower(CountryName))

# join tables and add missing country code
exchange_rates_grouped <- exchange_rates_grouped %>%
  left_join(
    country_dimension_table %>% select(CountryCode, CountryName) %>%
    mutate(CountryName = tolower(CountryName)),
    by = "CountryName")

# drop CountryCode with NAs, Rename second CountryCode column
exchange_rates_grouped <- exchange_rates_grouped %>%
  select(-CountryCode.x) %>%  
  rename(CountryCode = CountryCode.y) 

exchange_rates_grouped
```

```{r}
# check which country names did not match and count - the ones that could not be matched manually
empty_country_codes <- exchange_rates_grouped %>%
  filter(is.na(CountryCode) | CountryCode == "") %>% 
  select(CountryName)  %>% 
  group_by(CountryName) %>%
  summarize(EmptyEntriesCount = n())
empty_country_codes  
```

```{r}
# and primary key to dimension_table
fx_dimension_table <- exchange_rates_grouped %>%
  mutate(
  ID = ifelse(is.na(CountryCode), paste("NA", row_number(), Calendar.Year, sep = "_"), paste(CountryCode, Calendar.Year, sep = "_")))
fx_dimension_table
```

## 3.0 Making Fact Tables - Subsets (Trial)


The smoking prevalence, life expectancy and probability of death tables will be combined to one fact table first. The cost table will be joined at the end as it doesn't have gender. For the trial, a combination that has a value in each table has been chosen. In this case, continent in Africa, ZWE as country and both sexes as gender. This way we can check that the values from each table connects properly. As all the values should be in one row for this specific combination. 

```{r}
# Creating the subset for the specific combination (Contient = AFR, Country = ZWE, Gender = BTSX)
smoking_prev_3_subset <- smoking_prev_3 %>% filter((ContinentCode == "AFR") & (CountryCode == "ZWE") & (GenderCode == "BTSX") & (Year == 2010))
life_expectancy_3_subset <- life_expectancy_3 %>% filter((ContinentCode == "AFR") & (CountryCode == "ZWE") & (GenderCode == "BTSX") & (Year == 2010))
death_3_subset <- death_3 %>% filter((ContinentCode == "AFR") & (CountryCode == "ZWE") & (GenderCode == "BTSX") & (Year == 2010))
```

```{r}
smoking_prev_3_subset
life_expectancy_3_subset
death_3_subset
```

```{r}
trial_2 <- full_join(smoking_prev_3_subset, life_expectancy_3_subset, by = c("ContinentCode", "CountryCode", "Year", "GenderCode"))
trial_3 <- full_join(trial_2, death_3_subset, by = c("ContinentCode", "CountryCode", "Year", "GenderCode"))
trial_3

## all the values are in one row - confirming that it works
```

## 3.1 Joining Tables - Smoking + Probability of death + Life Expectancy

```{r}
smoke_lifexpect <- full_join(smoking_prev_3, life_expectancy_3, by = c("ContinentCode", "CountryCode", "Year", "GenderCode"))
smoke_lifexpect_death <- full_join(smoke_lifexpect, death_3, by = c("ContinentCode", "CountryCode", "Year", "GenderCode"))
smoke_lifexpect_death
```

## 4.0 Creating Final Fact table


### 4.1 Adding Primary key to cost table


A primary key will be made for each row through creating an "ID" based on the country code, year and gender code. For cost table, "B" will be added at the end of country code and year, so that all the costs will be in the both sex rows in the final fact table. 

```{r}
cost_cigs <- cost_3 %>%
  mutate(ID = paste(CountryCode, Year, "B"),
         ID = str_replace_all(ID, " ", "")) %>%
  select(ID, everything())

cost_cigs
```

### 4.2 Adding Primary key to taxes table


Primary keys are made in the same way as above.

```{r}
taxes_tobacco <- taxes_3 %>%
  mutate(ID = paste(CountryCode, Year, "B"),
         ID = str_replace_all(ID, " ", "")) %>%
  select(ID, everything())

taxes_tobacco
```

### 4.3 Adding Primary Key to Smoking Prevalence, Life Expectancy, Probablity of Death Table


Primary keys are made in the same way as above.

```{r}
smoke_lifexpect_death_wID <- smoke_lifexpect_death %>%
  mutate(ID = paste(CountryCode, Year, substr(GenderCode, 1, 1)),
         ID = str_replace_all(ID, " ", "")) %>%
  select(ID, everything())

smoke_lifexpect_death_wID #12567
```

### 4.4 Creating final Fact table


We will do a left join between smoking data and cost data, as there is no point looking into the cost of cigarettes if the information on smoking is not available. 

```{r}
fact_table_0 <- full_join(smoke_lifexpect_death_wID, cost_cigs, by = c("ID", "ContinentCode","CountryCode", "Year"))
head(fact_table_0)
nrow(fact_table_0)
```

```{r}
fact_table_1 <- left_join(fact_table_0, taxes_tobacco, by = c("ID", "ContinentCode","CountryCode", "Year"))
head(fact_table_1)
nrow(fact_table_1) 
```

```{r}
fact_table_1 %>% glimpse()
```

```{r}
fact_table_1 %>% filter(is.na(`GenderCode`))
# There were 12567 in the smoke_lifexpect_death_wID table and now there are 12642 when the cost has been joined.
# This means there are values in cost and taxes for certain country and year and both gender combinations that
# smoke_lifexpect_death_wID table didn't. As shown below there are 75 such rows, which makes sense (12642 - 12567).
# The gender code will need to be artificially put in for these. 
```

```{r}
# Artifically putting in the gender code for the extra cost and tax rows. 
fact_table_1$GenderCode[is.na(fact_table_1$GenderCode)] <- "BTSX"
fact_table_1 %>% filter(is.na(`GenderCode`))
```

```{r}
# R_Curr_mp is repeated in cost and taxes table, checking whether they have the same values - confirmed, they do. 
fact_table_1 %>% 
    filter(`R_Curr_mp.y` != `R_Curr_mp.x`) %>%
    select(-ContinentCode, -CountryCode, -Year, -GenderCode)
```

```{r}
# R_Price_mp_estimate is repeated in cost and taxes table, checking whether they have the same values - confirmed, they do. 
fact_table_1 %>% 
    filter(`R_Price_mp_estimate.x` != `R_Price_mp_estimate.y`) %>%
    select(-ContinentCode, -CountryCode, -Year, -GenderCode)
```

```{r}
# deselecting the repeated columns
fact_table_2 <- fact_table_1 %>% 
    select(-"R_Price_mp_estimate.x", -"R_Curr_mp.x") %>%
    rename("R_Price_mp_estimate" = "R_Price_mp_estimate.y", "R_Curr_mp" = "R_Curr_mp.y")
fact_table_2
```

```{r}
# unselecting columns that are not required
final_fact_table <- fact_table_2 %>%
    select(-M_Est_cig_curr_std, -M_Est_smk_curr_std, -WHOSIS_000015, -R_Curr_mp, -R_Sp_excise_estimate, -R_Ad_val_estimate, -R_VAT_estimate, -R_imp_duty_estimate, -R_Other_estimate, -ContinentCode) %>%
    rename(TobaccoPrevalence = M_Est_tob_curr_std, LifeExpectancy = WHOSIS_000001, Mortality = NCDMORT3070, LowestPricEstTobacco = R_Price_lowest_cost_estimate, PremiumPricEstTobacco = R_Price_premium_estimate, MediumPriceEstTobacco = R_Price_mp_estimate, TotalTax = R_total_tax_estimate)

final_fact_table
```

```{r}
final_fact_table %>%
    filter(CountryCode == "NZL" & Year == 2010) %>%
    select(-LowestPricEstTobacco, -PremiumPricEstTobacco)
```

## 5.0 Saving all the tables

```{r}
# write.csv(final_fact_table, file = "fact-table.csv", row.names = FALSE)
```

```{r}
# write.csv(country_dimension_table, file = "country_dimension_table.csv", row.names = FALSE)
# write.csv(years_dimensional_table, file = "years_dimensional_table.csv", row.names = FALSE)
# write.csv(gender_dimension_table, file = "gender_dimension_table.csv", row.names = FALSE)
# write.csv(continent_dimensional_table, file = "continent_dimensional_table.csv", row.names = FALSE)
# write.csv(indicators, file = "indicators.csv", row.names = FALSE)
# write.csv(fx_dimension_table, file = "fx_dimension_table.csv", row.names = FALSE)
```

## 6.0 Metadata table


Metadata table created to understand how many years are covered by each country for each variable. 

```{r}
# Filter the original data frame to keep only the selected columns and the CountryCode
filtered_df <- final_fact_table %>%
  select(-ID, -Year)

# Gather the selected columns to calculate counts for each country and each column
gathered_df <- filtered_df %>%
  gather(key = "Column_Name", value = "Value", -CountryCode, -GenderCode, na.rm = TRUE)

# Count the occurrences for each CountryCode, Column_Name pair
counted_df <- gathered_df %>%
  group_by(CountryCode, GenderCode, Column_Name) %>%
  summarise(Count = n())

# Spread the data to have column names as columns and counts as values
metadata_df <- counted_df %>%
  spread(key = Column_Name, value = Count, fill = 0)

metadata_df 
```

```{r}
# write.csv(metadata_df, file = "metadata.csv", row.names = FALSE)
```

## 7.0 Joining NZ data 


### 7.1 Trial with subset


The data for New Zealand has been specifically obtained from StatsNZ and MOH to analyse data around smoking in New Zealand. This data has been wrangled in julia and saved as "nz_data". Now we will load this up and combine it with the fact table. 

```{r}
nz_data <- read.csv("1. nz_data.csv")
```

```{r}
nz_data_final <- nz_data %>% 
    mutate(ID = paste(CountryCode, Year, "B"),
         ID = str_replace_all(ID, " ", "")) %>%
    select(ID, CountryCode, Year, GenderCode, everything())
```

```{r}
nz_data_final
```

The WHO NZL data will be merged with the data from nz_data. nz_data will override the WHO data where available, otherwise WHO data will remain.

```{r}
nz_data_who <- final_fact_table %>%
    filter(CountryCode == "NZL")
nz_data_who 
```

```{r}
nz_data_whole <- full_join(nz_data_who, nz_data_final, by = c("ID", "CountryCode", "GenderCode","Year"))
nz_data_whole %>% head()
```

```{r}
nz_data_whole_2 <- nz_data_whole %>%
    rename(TobaccoPrevalence_who = TobaccoPrevalence, MediumPriceEstTobacco_who = MediumPriceEstTobacco) %>%
    mutate(TobaccoPrevalence = ifelse(is.na(AveragePrev), TobaccoPrevalence_who, AveragePrev),
          MediumPriceEstTobacco = ifelse(is.na(cost_mean), MediumPriceEstTobacco_who, cost_mean)) %>%
    select(-cost_mean, -AveragePrev)
```

Removing the NZL data from full fact table and stacking the NZ data onto it. 

```{r}
final_fact_table_w_NZ <- final_fact_table %>%
    filter(CountryCode != "NZL")
```

```{r}
final_fact_table_updated <- bind_rows(final_fact_table_w_NZ, nz_data_whole_2)
final_fact_table_updated
```

```{r}
# write.csv(final_fact_table_updated, file = "fact-table.csv", row.names = FALSE)
```
